First Order Alternation

### 2018-07-10 quantifiers/draft.tex
ok

A \emph{first-order alternating automaton} (simply called automaton in
the following) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$,
where $\Sigma$ is the finite event alphabet, $X$ is a finite set of
variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
sentence, $F \subseteq Q$ is a set of \emph{final predicates} and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event, $X \cap
\set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive formula
such that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$
and each predicate that occurs in $\psi$ belongs also to $Q$. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}} \len{\psi}$.

@@

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I, q \in Q}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$. Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n) \in \Sigma[X]^*$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
such that:
\begin{compactitem}
\item the roots of $T_1, T_2, \ldots$ form a cube from
  $\cube{\minsem{\iota}}$, % --- because $\iota$ is satisfiable, such
                           % a cube always exists,
%
% \item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
                  % by $\eta$ to variables outside $X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
                  % $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
An execution $\F$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have:
  \[\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

@@

@@

Given a positive formula $\phi$, we define the \emph{dual} formula
$\dual{\phi}$ recursively on the structure of $\phi$ as follows: 
\[\begin{array}{rclcrclcrcl}
\dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} &&
\dual{(t \teq s)} & = & \neg(t \teq s) \\
\dual{(\exists x ~.~ \phi_1)} & = & \forall x ~.~ \dual{\phi_1} && 
\dual{(\forall x ~.~ \phi_1)} & = & \exists x ~.~ \dual{\phi_1} && 
\dual{(\neg(t \teq s))} & = & t \teq s \\
\dual{(q(x_1,\ldots,x_{\#(q)}))} & = & q(x_1,\ldots,x_{\#(q)})
\end{array}\]
Observe that, because predicate atoms do not occur negated in $\phi$,
there is no need to define dualization for formulae of the form $\neg
q(x_1,\ldots,x_{\#(q)})$.

@@

\begin{theorem}\label{thm:closure}
  Given automata $\A_i = \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, the following hold: 
  \begin{compactitem}
    \item\label{it1:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
      $\A_\cap = \tuple{\Sigma,X,Q_1 \cup Q_2, \iota_1 \wedge \iota_2,
      F_1 \cup F_2, \Delta_1 \cup \Delta_2}$, 
      %
    \item\label{it2:closure} $\lang{\overline{\A_i}} = \Sigma[X]^* \setminus
      \lang{\A_i}$, where $\overline{\A_i} =
      \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
        F_i,\overline{\Delta_i}}$ and, for all
      $i=1,2$: \[q(x_1,\ldots,x_{\#(q)}) \arrow{a(X)}{} \phi \in
      \Delta_i \iff q(x_1,\ldots,x_{\#(q)}) \arrow{a(X)}{} \dual{\phi}
      \in \overline{\Delta_i} \enspace.\]
  \end{compactitem}
  Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and
  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for all $i=1,2$.
\end{theorem}
\proof{\qed}

>>>>

A \emph{first-order alternating automaton} (simply called automaton in
the following) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$,
where $\Sigma$ is the finite event alphabet, $X$ is a finite set of
variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
sentence, $F \subseteq Q$ is the set of \emph{final} predicates and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event, $X \cap
\set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive formula
such that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$
and each predicate that occurs in $\psi$ belongs also to $Q$. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I, q \in Q}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$. Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
\Sigma[X]^*$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
such that:
\begin{compactitem}
\item the roots of $T_1, T_2, \ldots$ form a cube from
  $\cube{\minsem{\iota}}$, % --- because $\iota$ is satisfiable, such
                           % a cube always exists,
%
% \item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
                  % by $\eta$ to variables outside $X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
                  % $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
An execution $\F$ over $w$ is said to be \emph{accepting} if and only
if \begin{inparaenum}[(i)]
\item all paths in $\F$ have the same length $n$, and
\item the frontier of each tree $T \in \F$ is labeled with \emph{final
  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$ is a
  final predicate.
\end{inparaenum} In this case, we say that $\A$ accepts $w$ and
denote by $\lang{\A}$ the set of words accepted by $\A$.

In this paper, we answer the following questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

@@

Next, we give a logical characterization of acceptance, relative to a
given sequence of input events.

\begin{lemma}\label{lemma:acceptance}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and a
  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$, the
  formula $\accform{\alpha} = \pathform{\alpha} \wedge \bigwedge_{q
    \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$ is satisfiable
  if and only if $\A$ accepts a word $w$ such that $w_\Sigma =
  \alpha$.
\end{lemma}
\proof{ ``$\Rightarrow$'' Let $\I$ be an interpretation and $\nu$ be a
  valuation such that $\I,\nu \models \accform{\alpha}$. W.l.o.g. we
  consider $\I$ to be a minimal interpretation with this property. Let
  $w$ be the unique word such that $w_\Sigma = \alpha$ and $w_\Data =
  \nu$. By Lemma \ref{lemma:path-formula}, $\A$ has an execution $\F$
  over $w$ such that $\I = \I_{\F}$. To prove that $\F$ is accepting,
  we show that \begin{inparaenum}[(i)]
    \item\label{it1:acceptance} all paths in $\F$ have length $n$ and
      that
    \item\label{it2:acceptance} the frontier of $\F$ is labeled with
      final configurations only. \end{inparaenum} First, assume that
  (\ref{it1:acceptance}) there exists a path in $\F$ of length $0 \leq
  m < n$. Then there exists a node on the $m$-th level, labeled with
  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
  children. By the definition of the execution semantics of $\A$, we
  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta$ stands
  for $\nu[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
  $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
  q(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{m+1}$, thus
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
  contradicts the fact that $\I=\I_{\F}$ and that
  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\F$. Second, assume that
  (\ref{it2:acceptance}), there exists a frontier node of $\F$ labeled
  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
  \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$, by a
  similar reasoning as in the above case, we obtain that
  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.

  ``$\Leftarrow$'' Let $\F$ be an accepting execution of $\A$ over
  $w$, where $w_\Sigma = \alpha$. We prove that $\I_\F,w_\Data \models
  \accform{\alpha}$. By Lemma \ref{lemma:path-formula}, we obtain
  $\I_\F,w_\Data \models \pathform{\alpha}$. Since every path in $\F$
  is of length $n$ and all nodes on the $n$-th level of $\F$ are
  labeled by final configurations, we obtain that $\I_F,w_\Data
  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$,
  trivially.  \qed}

@@

Given a positive formula $\phi$, we define the \emph{dual} formula
$\dual{\phi}$ recursively as follows:
\[\begin{array}{rclcrclcrcl}
\dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} &&
\dual{(t \teq s)} & = & \neg(t \teq s) \\
\dual{(\exists x ~.~ \phi_1)} & = & \forall x ~.~ \dual{\phi_1} && 
\dual{(\forall x ~.~ \phi_1)} & = & \exists x ~.~ \dual{\phi_1} && 
\dual{(\neg(t \teq s))} & = & t \teq s \\
\dual{(q(x_1,\ldots,x_{\#(q)}))} & = & q(x_1,\ldots,x_{\#(q)})
\end{array}\]
Observe that, because predicate atoms do not occur negated in $\phi$,
there is no need to define dualization for formulae of the form $\neg
q(x_1,\ldots,x_{\#(q)})$.

@@

\begin{theorem}\label{thm:closure}
  Given automata $\A_i = \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, the following hold: 
  \begin{compactitem}
    \item\label{it1:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
      $\A_\cap = \tuple{\Sigma,X,Q_1 \cup Q_2, \iota_1 \wedge \iota_2,
      F_1 \cup F_2, \Delta_1 \cup \Delta_2}$, 
      %
    \item\label{it2:closure} $\lang{\overline{\A_i}} = \Sigma[X]^*
      \setminus \lang{\A_i}$, where $\overline{\A_i} =
      \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
        F_i,\dual{\Delta}_i}$ and, for $i=1,2$: \[\dual{\Delta}_i =
      \set{q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \dual{\psi} \mid
        q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi \in
        \Delta_i}\enspace.\]
  \end{compactitem}
  Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and
  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for all $i=1,2$.
\end{theorem}
\proof{\qed}

\section{The Emptiness Problem}

The problem of checking emptiness of a given automata is undecidable,
for automata using predicates of arity two or more, even if the theory
in which the (right-hand sides of the) transition rules are written
consists of only equalities and disequalities between variables, with
no quantifiers \cite{FarzanKincaidPodelski15}. Thus, we resort to an
abstraction-refinement loop that looks for an accepted word, by firing
longer and longer sequences of input events.

Since, for a countable data domain $\Data$, the set of finite words
$\Sigma[X]^*$ over a finite set of variables $X$ is countable, the
emptiness problem is recursively enumerable in this case. As a
consequence of Lemma \ref{lemma:acceptance}, if the given automaton is
not empty, a semi-algorithm based on a systematic exploration of the
sequences of input events will eventually discover an accepted word,
provided that the first-order data theory is decidable.

However, if the automaton is empty, the enumeration of input event
sequences will fail to terminate, in general. The classical way to
fight this divergence problem is to consider a \emph{coverage} partial
order, which is a subset of the following relation:

\begin{definition}\label{def:coverage}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and
  $S_1, S_2 \subseteq \bigcup_{q \in Q} \set{q(d_1,\ldots,d_{\#(q)})
    \mid d_1,\ldots,d_{\#(q)} \in \Data}$, we have $S_1 \sqsubseteq
  S_2$ if and only if 
  \begin{equation}\label{eq:po}
    \text{
      \begin{minipage}{0.7\textwidth}
        $\A$ accepts a word $w$ when started in a set of cubes $C_1
        \in S_1$ only if $\A$ accepts $w$ when started in a set of
        cubes $C_2 \in S_2$.
    \end{minipage}} 
  \end{equation}
\end{definition}
Since, in general the $\sqsubseteq$ relation is not definable using
available decidable (SMT) theories, we consider the following stronger
relation: $S_1 \supsetforex S_2$ if and only if, for each set of cubes
$C_1 \in S_1$, there exists a set of cubes $C_2 \in S_2$ such that
$C_1 \supseteq C_2$.

\begin{lemma}\label{lemma:coverage}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  sets of cubes $S_1, S_2 \subseteq \bigcup_{q \in Q}
  \set{q(d_1,\ldots,d_{\#(q)}) \mid d_1,\ldots,d_{\#(q)} \in \Data}$
  such that $S_1 \supsetforex S_2$, we have $S_1 \sqsubseteq S_2$.
\end{lemma}
\proof{  
\qed}
